Leaf mass per area, not total leaf area, drives differences in above-ground biomass distribution among woody plant functional types

### 2016-04-26 manuscript.Rnw
Abstract updated; reference added; minor edits

%The Summary for research papers, which must be usable as a stand-alone document, must not exceed 200 words and should be organized using four bullet points to indicate (1) the research conducted, including the rationale, (2) methods, (3) key results and (4) the main conclusion, including the key points of discussion.
\begin{itemize}
\item We tested the hypothesis that biomass distribution in leaves and stems is related to the leaf mass per area, an important leaf trait, across major woody plant functional types (PFTs) (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm).
\item We used a large compilation of plant biomass and size observations, including observations of \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species. We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, and climate on biomass distribution.
\item The ratio of leaf mass to above-ground woody mass ($\mfms$) varied strongly between PFTs. We found that $\mfms$, once corrected for plant height, was proportional to leaf mass per area across PFTs. As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass. Climate consistently explained very little additional variation in biomass distribution at a given plant size.
\item Combined, these results demonstrate consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain global vegetation models.
\end{itemize}

@@

The goal of the current work is to better understand differences among PFTS in their biomass distribution. Past work has shown that gymnosperms tend to hold a much larger leaf mass compared to angiosperms (at a given stem mass) \citep{oneilldeangelis1981,enquistniklas2002,reich2014,poorter2012,poorter2015}. The cause of these differences are not yet fully understood, but could arise from either differences leaf mass per area (LMA) \citep{poorter2009}, or the amount of leaf area maintained \citep{chabothicks1982,bond1989}, or both. It is also unknown whether there exist differences among finer-scale groupings, for example between evergreen and deciduous angiosperms. While past analyses have often spanned many sites and / or species, researchers have been unable to further investigate the factors underlying observed differences in biomass distribution among PFTs, because available data only included a limited number of variables.

Here, we use the recently-compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which includes observations of biomass and other size metric for \Sexpr{nrow(baad_all$data)} individual plants (Fig. 1), to establish the underlying cause of biomass distribution among PFTs, and also to test for systematic shifts in biomass distribution with climate. We focus on the distribution of biomass among leaf and above-ground woody tissues in individual plants. To quantify this distribution we analyse two related metrics: 1) the ratio of leaf mass to stem mass, $\mfms$; and 2) the ratio of leaf area to stem mass, $\afms$. To understand the factors driving variation in these quantities, we decompose these ratios into component variables, each of which can be quantified in our dataset. The ratio of leaf mass to stem mass, can decomposed as
\begin{equation}
\label{eq:mfmspart}
\frac{\mf}{\ms} =  \frac{\mf}{\af} \times \frac{\af}{\as} \times \frac{\as}{\ms},
\end{equation}
where $\mfaf$ is the LMA, $\afas$ the ratio of total plant leaf area to basal stem area, and $\asms$ is the ratio of stem basal area to above-ground woody mass. Similarly, the ratio of leaf area to stem mass can decomposed as
\begin{equation}
\label{eq:afmspart}
\frac{\af}{\ms} =  \frac{\af}{\as} \times \frac{\as}{\ms}.
\end{equation}
The terms $\mfaf$ and $\afas$ are themselves prominent traits known to vary among species \citep{wright2004,togashi2015}, suggesting a potential to explain differences in $\mfms$ among PFTs. The third term is known to vary strongly with individual size, decreasing as a plant grows. However, \citet{chave2014} found that a single model of $\ms$ incorporating $\as$, wood density, and $H$ to hold across diverse tropical trees, irrespective of vegetation type, or environmental factors.

@@

\begin{enumerate}
  \item Individual height captures the most majority of variation in $\mfms$ and $\afms$, reflecting the fact that height can vary several orders of magnitude within the lifespan of a single individual and thus represents the dominant form of variation in eq . \ref{eq:mfmspart}-\ref{eq:afmspart}.
  \item That these height-related effects on biomass distribution are concentrated in the term $\asms$, and that while $\asms$ may vary considerably among individuals, there are no systematic differences in $\asms$ among PFTs.
  \item That after controlling for the effects of individual height, differences in $\mfms$ among PFTs can arise due to differences in either, or both, of a) LMA, because PFTs tend to differ LMA \citep{poorter2009}, or b)$\afas$, because gymnosperms are hypothesised to maintain a greater leaf area than angiosperms \citep{chabothicks1982, bond1989}.
  \item That any differences in $\afas$ among PFTs also leads to differences in $\afms$.
\end{enumerate}

A final aim was test for significant shifts in biomass distribution with climate. Little is known about global-scale patterns in biomass distribution in relation to climate, but one might expect biomass distribution to shift with precipitation or mean annual temperature, as small-scale comparisons have shown such responses \citep{berninger1994,callaway1994,delucia2000,poyatos2007}. \citet{reich2014} has also shown that after controlling  for the large size-related variation in biomass distribution, the ratio of leaf to total plant mass (averaged across whole stands, either including or excluding below-ground components) increases with mean annual temperature in both gymnosperms and angiosperms. Thus, we aimed to test for climate-related variation in $\mfms$ and $\afms$, and their component terms with PFTS.

@@

To test for climate effects on biomass distribution, we estimated mean annual temperature (MAT), precipitation (MAP) and Potential evapotranspiration (PET) for each study location. MAT and MAP were extracted from Worldclim \citep{worldclim}, PET from the Global-PET database (http://www.cgiar-csi.org, \citet{zomer2008}), and an aridity index was calculated as PET/MAP. To assess the coverage of the global climate space by the dataset, we also extracted MAT and MAP across the globe from the global land cover database GLC-SHARE \citep{lathamglobal},but excluding areas without woody vegetation.

@@

Using variance partitioning, we quantified the contribution of plant height, plant functional type, and climate to the overall variation in $\mfms$ and for all three component variables. The amount of variance explained by quantitative climate variables (MAP and MAT) was assessed with GAMs, with variables sequentially added to the model and the explained variance ($R^2$) calculated. We also compared this approach to another method using linear mixed-effects models instead of GAMs \citep[calculating the $R^2$ for linear mixed-effects models for the fixed effects only following][]{nakagawa2013}, as well as third method called ``independent effects analysis''. The three methods generally agreed on the ranking of variable importance (see Fig. S4 and S5).

>>>>

%The Summary for research papers, which must be usable as a stand-alone document, must not exceed 200 words and should be organized using four bullet points to indicate (1) the research conducted, including the rationale, (2) methods, (3) key results and (4) the main conclusion, including the key points of discussion.
\begin{itemize}
\item Here we aim to understand differences in biomass distribution between major woody plant functional types (PFTs) (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm) in terms of underlying traits, in particular the leaf mass per area, a key leaf trait.
\item We used a large compilation of plant biomass and size observations, including observations of \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species. We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, leaf mass per area, total leaf area, stem basal area and climate on above-ground biomass distribution. 
\item The ratio of leaf mass to above-ground woody mass ($\mfms$) varied strongly between PFTs. We found that $\mfms$, at a given plant height, was proportional to leaf mass per area across PFTs. As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass. Climate consistently explained very little additional variation in biomass distribution at a given plant size.
\item Combined, these results demonstrate consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain global vegetation models.
\end{itemize}

@@

The goal of the current work is to better understand differences among PFTs in their biomass distribution. Past work has shown that gymnosperms tend to hold a much larger leaf mass compared to angiosperms (at a given stem mass) \citep{oneilldeangelis1981,enquistniklas2002,reich2014,poorter2012,poorter2015}. The cause of this difference is not yet fully understood, but could arise from either differences leaf mass per area (LMA) \citep{poorter2009}, or the amount of leaf area maintained \citep{chabothicks1982,bond1989}, or both. It is also unknown whether there are differences among finer-scale groupings, for example between evergreen and deciduous angiosperms. While past analyses have often spanned many sites and / or species, researchers have been unable to further investigate the factors underlying observed differences in biomass distribution among PFTs, because available data only included a limited number of variables.

Here, we use the recently-compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which includes observations of biomass and other size metrics for \Sexpr{nrow(baad_all$data)} individual plants (Fig. 1), to establish the underlying cause of biomass distribution among PFTs, and also to test for systematic shifts in biomass distribution with climate. We focus on the distribution of biomass among leaf and above-ground woody tissues in individual plants. To quantify this distribution we analyse two related metrics: 1) the ratio of leaf mass to stem mass, $\mfms$; and 2) the ratio of leaf area to stem mass, $\afms$. To understand the factors driving variation in these quantities, we decompose these ratios into component variables, each of which can be quantified with our dataset. The ratio of leaf mass to stem mass, can be decomposed as
\begin{equation}
\label{eq:mfmspart}
\frac{\mf}{\ms} =  \frac{\mf}{\af} \times \frac{\af}{\as} \times \frac{\as}{\ms},
\end{equation}
where $\mfaf$ is the LMA, $\afas$ the ratio of total plant leaf area to basal stem area, and $\asms$ is the ratio of stem basal area to above-ground woody mass. Similarly, the ratio of leaf area to stem mass can decomposed as
\begin{equation}
\label{eq:afmspart}
\frac{\af}{\ms} =  \frac{\af}{\as} \times \frac{\as}{\ms}.
\end{equation}
The terms $\mfaf$ and $\afas$ are themselves prominent traits known to vary among species \citep{wright2004,togashi2015}, suggesting a potential to explain differences in $\mfms$ among PFTs. The third term is known to vary strongly with individual size, decreasing as a plant grows. However, \citet{chave2014} found that a single model of $\ms$ incorporating $\as$, wood density, and $H$ to hold across diverse tropical trees, irrespective of vegetation type, or environmental factors.

@@

\begin{enumerate}
  \item Individual height captures the most majority of variation in $\mfms$ and $\afms$, reflecting the fact that height can vary several orders of magnitude within the lifespan of a single individual and thus represents the dominant form of variation in Eqns \ref{eq:mfmspart}-\ref{eq:afmspart}.
  \item Height-related effects on biomass distribution are concentrated in the term $\asms$, and that while $\asms$ may vary considerably among individuals, there are no systematic differences in $\asms$ among PFTs.
  \item After controlling for the effects of individual height, differences in $\mfms$ among PFTs can arise due to differences in either, or both, of a) LMA, because PFTs tend to differ LMA \citep{poorter2009}, or b)$\afas$, because gymnosperms are hypothesised to maintain a greater leaf area than angiosperms \citep{chabothicks1982, bond1989}.
  \item That any differences in $\afas$ among PFTs also lead to differences in $\afms$.
\end{enumerate}

A final aim was to test for shifts in biomass distribution with climate. Little is known about global-scale patterns in biomass distribution in relation to climate, but one might expect biomass distribution to shift with precipitation or mean annual temperature, as small-scale comparisons have shown such responses \citep{berninger1994,callaway1994,delucia2000,poyatos2007}. \citet{reich2014} has also shown that after controlling  for the large size-related variation in biomass distribution, the ratio of leaf to total plant mass (averaged across whole stands, either including or excluding below-ground components) increases with mean annual temperature in both gymnosperms and angiosperms. Thus, we aimed to test for climate-related variation in $\mfms$ and $\afms$, and their component terms with PFTs.

@@

To test for climate effects on biomass distribution, we estimated mean annual temperature (MAT), precipitation (MAP) and Potential evapotranspiration (PET) for each study location. MAT and MAP were extracted from Worldclim \citep{worldclim}, PET from the Global-PET database (http://www.cgiar-csi.org, \citet{zomer2008}), and an aridity index was calculated as PET/MAP. To assess the coverage of the global climate space by the dataset, we also extracted MAT and MAP from Worldclim for each 0.5\degree cell across the globe but excluding areas without woody vegetation, as assessed from the global land cover database GLC-SHARE \citep{lathamglobal}.

@@

Using variance partitioning, we quantified the contribution of plant height, plant functional type, and climate to the overall variation in $\mfms$ and for all three component variables. The amount of variance explained by quantitative climate variables (MAP and MAT) was assessed with GAMs, with variables sequentially added to the model and the explained variance ($R^2$) calculated. We also compared this approach to another method using linear mixed-effects models instead of GAMs \citep[calculating the $R^2$ for linear mixed-effects models for the fixed effects only following][]{nakagawa2013}, as well as third method called ``independent effects analysis''. The three methods generally agreed on the ranking of variable importance (see Figs. S4 and S5).
